---
title: "Challenge 2 Instructions"
author: "Sean Conway"
description: "Data wrangling: using group() and summarise()"
date: "12/21/2023"
format:
  html:
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
categories:
  - challenge_2
  - railroads
  - faostat
  - hotel_bookings
---

```{r}
#| label: setup
#| warning: false
#| message: false

knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

Today's challenge is to

1)  read in a data set, and describe the data using both words and any supporting information (e.g., tables, etc)
2)  provide summary statistics for different interesting groups within the data, and interpret those statistics

## Read in the Data

Read in one (or more) of the following data sets, using the correct R package and command.

-   railroad\*.csv or StateCounty2012.xls ⭐
-   FAOstat\*.csv or birds.csv ⭐⭐⭐
-   hotel_bookings.csv ⭐⭐⭐⭐

Add any comments or documentation as needed. More challenging data may require additional code chunks and documentation.

## Describe the data

Using a combination of words and results of R commands, can you provide a brief, high level description of the data? Describe as efficiently as possible where/how the data was (likely) gathered, indicate the cases and variables (both the interpretation and any details you deem useful to the reader to fully understand your chosen data).

## Provide Grouped Summary Statistics

Conduct some exploratory data analysis, using dplyr commands such as `group_by()`, `select()`, `filter()`, and `summarise()`. Find the central tendency (mean, median, mode) and dispersion (standard deviation, mix/max/quantile) for different subgroups within the data set. Describe what you find.

```{r}

```

